---
title: "R Notebook"
output: html_notebook
---

# Summary

Hierarchical (HM) is a generalization of regression modeling. HM is also known as multilevel model or random effects model. HM considers dependencies of observations within groups. Why considering the within group dependencies?

If we opt to pool all samples together and fit a fixed effect model in cases where there are well-defined hierarchical structures, we won't be able to make group-specific predictions.
If we opt to estimate group-specific effects by fitting a fixed effect model, we are risky of overfitting in groups of few samples, and consequences poor out-of-sample predictions. 

In random effect model, we still estimate group-specific parameters, but these parameters are further modeled by hyperparameters. As a result, the group-specific parameter estimand are results of two sources: (1), data within the group; (2), data of other groups also play a role. 

# Case study using radon contamination data

```{r}
library(rstan)
load("~/Git/stats/mixed/radon.rdt")
attach(radon)

hist(log.radon)
table(basement)
hist(uranium)
unique(county)
```

### complete pool

```{r}
pool = stan_model("~/Git/stats/mixed/pool.stan")

x <- cbind(1, radon$basement)
data <- list(N = dim(x)[1], D = dim(x)[2], x = x, y = radon$log.radon)

fit.pool <- sampling(pool, data = data)
w <- extract(fit.pool)$w

plot(basement, log.radon); lines(basement, x %*% apply(w, 2, mean), col='red')
```

### mixed effect model (partial pooling)

```{r}
mixed = stan_model("~/GitHub/stats/mixed/mixed.stan")

data$C = length(unique(county))
data$l = county
fit.mixed <- sampling(mixed, data = data)
w <- extract(fit.pool)$w

```

> how to use HLM results to make predictions on a new group
> likely expected mean of the group-specific parameters

> what is the expected mean of the group-specific parameters?
> it deviates a bit, and is always between the overall expected mean and the within group expected mean. 

```{r}
print(fit.mixed, pars = "mu_w")
print(fit.mixed, pars = c("w[2,1]", "w[2,2]"))

radon.1 =  radon[radon$county == 2, ]
summary(lm(log.radon ~ basement, data = radon.1))

```
