---
title: "Statistical Models of Meta-analysis"
output: html_document
---

## Meta-analysis of 9 randomized controlled trials (RCTs) of breast cancer screening.

```{r}
library(rstan)

rct <- data.frame(study = 1:9) # study id
rct$d1 <- c(218, 87, 126, 135, 105, 107, 66, 88, 105) # number of death in group 1
rct$n1 <- c(31000, 20695, 38589, 38491, 25214, 19711, 40318, 21650, 53884) # number of total in group 1
rct$d0 <- c(262, 108, 104, 173, 108, 105, 45, 162, 251) # number of death in group 0
rct$n0 <- c(31000, 20783, 18582, 37403, 25216, 19694, 19943, 29961, 106956) # number of total in group 0

rct$p1 <- rct$d1/rct$n1 # risk in group 1
rct$p0 <- rct$d0/rct$n0 # risk in group 0
rct$rr <- rct$p1/rct$p0 # relative risk of group 2 to group 1
rct$lrr <- log(rct$rr) # log scale
rct$lse <- sqrt((1 - rct$p1)/(rct$p1 * rct$n1) + (1 - rct$p0)/(rct$p0 * rct$n0)) # se using delta method
rct$lower <- exp(rct$lrr - qnorm(.975) * rct$lse)
rct$upper <- exp(rct$lrr + qnorm(.975) * rct$lse)

library(metafor)
p <- forest(x = rct$rr, ci.lb = rct$lower, ci.ub = rct$upper, slab = paste("Study", rct$study), refline = 1)
text(min(p$xlim), .88 * max(p$ylim), "Study", pos = 4, font = 2)
text(max(p$xlim), .88 * max(p$ylim), "Relative Risk [95% CI]", pos = 2, font = 2)
```

## Meta-analysis models are in fact hiearchical model in Bayesian framework

$$y_i \sim N(\theta_j, \sigma_j^2)$$
$$\theta_j \sim N(\mu, \tau^2)$$
Equivalently:

$$y_i = \mu + u_j + e_j$$
$$u_j \sim N(0, \tau^2)$$
$$e_j \sim N(0, \sigma^2)$$

$e$ and $u$ are two variance components that capture within study sampling error and between study sampling error, respectively.

## There are, in general, 3 ways to estimate the random effects

1. No pooling: there is a separate model for each study, so that $\theta_j$ = $y_j$. This is a equivalent to set $\tau$ as $\infty$.
2. Complete pooling: patients in each study are random samples from a common distribution, so that $\theta_j$ = $\mu$. This is a equivalent to set $\tau$ as 0. This is also equivalent to a fixed-effect model.
3. Partial-pooling: In this case, $\tau$ is unknown and is due to be estimated.

```{r}
data = list(J = 9, y = rct$lrr, sigma = rct$lse)

fixed.stan = stan_model("~/Git/stats/meta/fixed.stan")

fixed.fit = sampling(fixed.stan, data)
print(fixed.fit)

# consistent with fixed effect model using inverse-variance weighting method

sum(data$y * data$sigma^{-2}) / sum(data$sigma^{-2})
sqrt(1 / sum(data$sigma^{-2}))

random.stan = stan_model("~/Git/stats/meta/random.stan")

random.fit = sampling(random.stan, data)
print(random.fit)
```
